---
title: "found R guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{found R guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  # `eval` set to FALSE if within CI since GHA runner does not have enough computational resources to render guide
  # @TODO: look into using self-hosted runners depending on future hardware access
  eval = Sys.getenv("CI") == "",
  collapse = TRUE,
  comment = "#>"
)
```

first we must ensure that all vignette dependencies are installed and loaded:

```{r}
library(found)

if (!(requireNamespace("pak", quietly = TRUE))) {
  install.packages("pak")
}
deps <- list(
  "Seurat" = "Seurat",
  "SeuratData" = "satijalab/seurat-data",
  "irlba" = "irlba"
)
mapply(function(name, pkg) {
  if (!(requireNamespace(name, quietly = TRUE))) {
    pak::pak(pkg)
  }
}, names(deps), deps) |> invisible()

suppressPackageStartupMessages({
  library(irlba)
  library(SeuratObject)
  library(Seurat)
  library(SeuratData)
})
```

for this example vignette, we will use the `ifnb` dataset provided by the `SeuratData` package, which we can load as follows:

```{r}
if (!("ifnb" %in% InstalledData()[["Dataset"]])) {
  InstallData("ifnb")
}
# ignore the various messages generated by LoadData
so <- LoadData("ifnb") |>
  suppressWarnings() |>
  suppressMessages()
```

we can then utilize existing `found` pipelines to run label refinement:

```{r}
out <- HiDDEN(
  so, "stim", "CTRL",
  adapters$Pipeline(methods$run_lognorm_pca, methods$reg_logit, methods$bin_kmeans),
  # ⚠️: X must be observation by feature format, so we transpose the output of LayerData!
  X = t(LayerData(so, "counts")), k = 15L
)
```

however, using the `to_step` wrapper, we can also inject (almost) any R function into various steps of the pipeline!
for example, below we switch to using the pearson residual based transform implemented by `sctransform` as a VST and `irlba` for dimensionality reduction.
given that our dimensionality reduction function is now written in R, we can also change the pipeline dependencies to take the entire seurat object as input, instead of having to extract the expression matrix.

```{r}
so <- SCTransform(so, verbose = FALSE) |> suppressWarnings()
sct_pca <- function(so, k) {
  prcomp_irlba(
    t(LayerData(so, assay = "SCT", layer = "scale.data")),
    n = k
  )[["x"]]
}
out <- HiDDEN(
  so, "stim", "CTRL",
  adapters$Pipeline(to_step(sct_pca), methods$reg_logit, methods$bin_kmeans),
  so = so, k = 15L
)
```

we can also try utilizing the automatic hyperparameter tuning implemented in `found`:

```{r}
outs <- HiDDENt(
  so, "stim", "CTRL",
  adapters$Pipeline(to_step(sct_pca), methods$reg_logit, methods$bin_kmeans),
  tune$FixPointTuner(3, 8, 0.02),
  so = so
)
# we can see what the automatically selected hyperparameter `k` is:
outs[["chosen"]]
```
